Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-03-09 analysis/paper/paper.qmd
draft editing

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

keywords: |
  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
highlights: |
  Indigenous community in northeastern Taiwan engaged in regional deer trade were identified from faunal remains.
  The high freqiencies of deer bones at Kiwulan reveals continously cultural persistence in deer consumption.
  Changes in deer butchery patterns and cooking practices indicates an intensified hunting during the European colonization period.
  Increased reliance on suids during the European period suggests adaptive strategies to sustain local subsistence.
---

@@

# Materials and Methods
The faunal assemblages from Kiwulan analyzed in this study were sampled from 40 adjacent units at the residential area excavated during the first season (2001-2003) [@Chen2007]. These 40 units were selected because of their location that were believed to be the major residential area with thick archaeological deposits and relatively intact archaeological contexts without disruption from modern construction. The spatial continuity can provides a consistent framework for understanding food practices over time on a settlement scale. Also, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences. I mainly focus on vertebrate species due to their economic values in the trade market. The contexts of faunal remains include dwelling spaces and middens that were associated with daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.

To explore changes in faunal remains over time, the original arbitrary levels of Kiwulan were assigned into six natural strata spanning from the 14th century to the 19th century, referred as cultural layers (CL1-Cl6) in this study. Each stratum was determined according to documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the specific time of European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence. Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers without attaching to cranium were excluded since they were less likely related to dietary practices. To obtain generalized pattern of animal use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumtion of that taxa [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from MNI that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of remains by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site scale with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison across animal classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange and in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded to reduce overrepresentation due to their much higher preservation rate and quantities compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as additional source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease [@Lupo1997], which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.

Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are characteristic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

The data and R code used for these analyses and visualizations are openly available at xxx for research reproducibility and transparency (@Marwick2017; @MarwickandWang2022).

@@

The faunal assemblage analyzed in this study from Kiwulan is composed of `r total_NISP` bone fragments based on NISP (@tbl-counts). Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`). To further compare general patterns of animal consumption in terms of their economic values over time, animals in the same family with similar body size were grouped together. Thus, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were group into the category, bovine. Similarly, domestic pig (n = 2, from CL4 and CL6) was integrated into the suid category. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet. In general, the faunal remains show the frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A). A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C). Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals. The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site. Within those identifiable mammals, deer is the main animal from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by bovine (@fig-NISP-proportion: D). Suid was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2. In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where suid and fish have slightly higher proportions compared to those in other layers. Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model. The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which suggests little bias from potential sample size.

@@

@tbl-ubiquity shows deer had been the most widespread animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), and lowest in CL2. The second and third animal regularly used are suid and muntjac. Although those two animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different. Suid was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than suid. Bovine had a low ubiquity in CL5 and high ubiquity in CL6. Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3. The result of animal ubiquity correspond with their frequencies (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of suid and muntjac in CL4.

Although deer was predominant throughout all temporal sequences, variations in their skeletal parts could be identified. Deer bones were grouped into six body portions [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those parts should be transported together [e.g., @Stiner1991]. The generally low frequency of axial elements might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods. Those differences might indicate changes in processing decisions of meat portions of deer.

Considering the impacts of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of bones. For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 are less fragmented (index = `r frag_CL3_CL4[2]`). However, the fragmentation rates are not significantly differed between layers.

@@

Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. @fig-deer-cutmark shows that at least 50 % of joints in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proportion of joints with cut marks suggests intensive processing of deer throughout the site's occupation. In addition, cut marks on lower limbs should be associated with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1). Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing. In addition, natural alterations to bones resulting from post-deposition impacts were also observed, such as weathering and root etching [@Gifford2018natural], on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify). Those post-depositional impacts might have led to an underestimation of cut marks in general.

@@

<!--skeletal part frequencies and butchery patterns of deer-->
Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer. Changes in cooking practices are also evident. The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997]. The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices. This could be further confirmed through organic residue analysis on pottery. Other meat preparation methods, such as drying through salting and smoking, are also possible [e.g., @Spyrou2019]. Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963]. The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.

>>>>

---
title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Institute of History and Philology, Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

keywords: |
  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
highlights: |
 Faunal analysis from Kiwulan suggests Indigenous participation in the European long-distance trade network. Evidence indicates the continued cultural persistence of deer consumption. Shifts in deer butchery patterns and cooking practices suggest intensified hunting during the European colonization period. The increased reliance on suids reflects adaptive strategies for sustaining local subsistence.

---

@@

# Materials and Methods
The faunal assemblages analyzed in this study were sampled from 40 adjacent units in the residential area excavated during the first season (2001-2003) [@Chen2007]. These 40 units were selected because they represent the main residential area, with thick archaeological deposits and relatively intact contexts without modern disturbance. The spatial continuity of these units provides a consistent framework for understanding food practices over time at the settlement scale. Additionally, the use of the same excavation and recovery methods allows for comparison between assemblages from different temporal sequences. I primarily focus on vertebrate species due to their economic importance in local subsistence and trade. The faunal remains analyzed were from dwelling spaces and middens, as these are associated with daily subsistence practices, while remains from burials are excluded due to their potential ritual purposes rather than daily practices.

To explore changes in faunal remains over time, the original arbitrary excavation levels (10 cm) at Kiwulan were assigned into six natural strata or cultural layers (CL1-CL6), spanning from the 14th century to the 19th century. Each stratum was determined based on documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. The diagnostic materials include pipes and stoneware commonly found at sites associated with European expansion in Southeast Asia, such as Anping jars and martavan  [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], and clay bricks associated with Chinese migrants [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum corresponding to the specific time of European arrival is challenging due to its brief duration, the strata separated based on the criteria above still provide a relatively robust framework for comparing archaeological records before and after foreign presence. Among these six temporal layers, the fourth cultural layer (CL4) is associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, linked to the large influx of Chinese migrants in Yilan.

Only faunal assemblages identified to at least the family level were considered in this study to ensure consistent comparison across time sequences. Deer antlers detached from the cranium were excluded because they were less likely to be related to dietary practices. To obtain a generalized pattern of animal consumption, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumption of that taxa [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units containing a given taxon to the total number of units per cultural layer. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012]. Unlike MNI, which determines the actual numbers of individual animals recovered from remains, NISP offers a broader overview by considering all fragmented bone assemblages. Despite criticism for potentially overestimating the number of animals, NISP better captures variation in species abundance with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison across animal classes, the normalized NISP were used to avoid biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin could be a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This analysis can provide insights into how Indigenous people utilized deer in response to possible high economic demand when driven by their involvement in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty parts (e.g., upper limbs), mid-meaty parts (e.g., ribs), and less meaty parts (e.g., skulls and lower limbs) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded to avoid overrepresentation, given their much higher preservation rate and quantities compared to other parts. While lower limbs and skull bones contain minimal meat, they are often valued in many cultures for marrow and soft tissues, such as the tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as an additional source of nutrition in the later interpretation. Similarly, the degree of fragmentation could indicate the extraction of marrow or grease [@Lupo1997] and was calculated by dividing the Minimum Number of Elements (MNE) by NISP. In this approach, lower values represent a higher degree of fragmentation, and vice versa.

Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, along with natural marks like weathering and root etching to understand taphonomic effects. Cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time. Moreover, cut marks on lower limbs related to skinning processes, were examined and compared. Such skinning marks are characteristic of circular and straight incisions left by a series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

The data and R code used for these analyses and visualizations are openly available at https://github.com/LiYingWang/kwlenv for research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].

@@

The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts). Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`). To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pig (n = 2, from CL4 and CL6) were integrated into the suid category. Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source across all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C). Fish and birds are both found from CL5 to CL3; however, their proportions (< 10%) are much lower compared to mammals. Reptiles are represented by a single piece of turtle shell found in CL4, suggesting that they might not be a common food source at the site. Among the identifiable mammals, deer is the main animal from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated by bovine (@fig-NISP-proportion: D). Suid remains were present from CL6 to CL3, with a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5 and was continuously present until CL2. There is an increase in animal diversity in CL4 and CL5. Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied across layers, the relationship between the number of taxa and sample size (NISP) was examined by a linear regression model. The results show no correlation between animal diversity and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), suggesting minimal bias from sample size variation.

@@

@tbl-ubiquity shows deer was the most widely used animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), and the lowest in CL2. The second and third most commonly used animals are suid and muntjac. Although both reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, ,and `r CL4_ubi_munt` for muntjac), their presence patterns differ slightly. Suid was utilized as early as CL6, declined in CL3, and absence in CL2, while muntjac showed a cultural lag, appearing one phase later than suid. Bovine had low ubiquity in CL5 and high ubiquity in CL1. Small animals (birds, fish, and turtle) had much lower ubiquity than large animals, with a slight increase in ubiquity in CL3. The result of ubiquity corresponds with the frequency data (NISP), suggesting that deer was the most common resource at the site, with an increased use of suid in CL4.

Although deer were predominant throughout all temporal sequences, variations in their skeletal parts can be identified. Deer bones were grouped into six body portions [@Madrigal2004], including the head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The results (@fig-deer-portion: A) show that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by the upper hindlimb and the less-meaty portions, lower limb and foot. The consistently higher frequencies of upper and lower limbs suggest that these parts were likely transported together [e.g., @Stiner1991]. The generally low frequency of axial elements may be due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern is observed in the frequency of cranial bones, which were scarce in CL6, increased to the highest number in CL4, and then decreased thereafter. This may reflect changes in deer processing decisions or transportation strategies over time.

Considering the potential impact of differential bone density on the preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested using Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer and calculated by dividing NISP by the expected NISP, then multiplying by 100%. The results (@fig-deer-portion: B) show no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias had little impact on the taphonomic processes of the bones. As for bone fragmentation, CL3 shows a higher degree of fragmentation (`r frag_CL3_CL4[1]`), while bones in CL4 are less fragmented (`r frag_CL3_CL4[2]`).

@@

Surface modifications, including cut marks, burning damage, and fractures, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Cut marks appeared on 50% of bones in CL2, 30% in earlier periods (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, indicating butchering activities related to disarticulation and meat removal. At least 50% of joints in CL4 presents cut marks as shown in @fig-deer-cutmark. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Direct comparisons across all layers are challenging since not all joints were found in CL6, CL3, and CL2. Nevertheless, the overall high frequency of joints with cut marks suggests intensive deer processing throughout the site's occupation. In addition, cut marks on lower limbs show traces associated with skinning (@fig-cutmark-img: e). Those cut marks were primarily found on metapodials in CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1). Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), accompanied by narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks, represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e) [@Bello2023], were observed on a few bones, suggesting the evidence of carnivore gnawing. Additionally, natural post-depositional alterations, such as weathering and root etching [@Gifford2018natural], were observed on approximately one-third of the bones across all layers (@fig-fracture-img: c, d; @tbl-deer-modify). These post-depositional impacts may have contributed to an underestimation of cut marks.

@@

<!--skeletal part frequencies and butchery patterns of deer-->
Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer. Changes in cooking practices are also evident. The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997]. The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices. This could be further confirmed through organic residue analysis on pottery. Other meat preparation methods, such as drying, are also possible [e.g., @Spyrou2019]. Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963]. Moreover, some features containing long, thin (~4 cm in diameter) or Y-shaped racks may have been associated with drying facilities for meat or hides [@Chen2007; @Friesen2001]. The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.
